### 2020-05-24 paper.tex
Rewrite Interviews

Informal interviews allowed us to gain first insights in how software engineers work with RP in their daily jobs. We talked to six professionals and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.

Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications. The fifth partner was the creator of rxfiddle, a visualization tool for RxJS \mintinline{TypeScript}{Observable}s and with this, proficient RxJS user. Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}-based, reactive API layer serving a web frontend application.

We identified several recurring topics throughout all the interviews. All of the subjects noted that they think of RP as an invaluable tool to describe and model dependencies between multiple data sources. Most of them appreciate how RP allows them to implement data transformations in a concise and composable way. The complexity of concurrent programming can be hidden well behind a simple DSL one subject noted further.

For the implementation, all subjects noted that a language with a strong type system is to be preferred: The compiler gives the guarantee of basic code correctness which is why e.g. TypeScript is used by all the subjects working with RxJS and Angular.

None of the subjects remembered debugging RP code to be a positive experience, independent of the technology used. All of the subjects use traditional debugging tools. Redux\footnote{\url{https://redux.js.org/}} DevTools \footnote{\url{https://github.com/reduxjs/redux-devtools}} is in use by all four subjects working with ngrx. There was no mention of any more specific debugging tool in regards of akka and akka-streams.

One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented. (ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{https://rxjs.dev/api/operators/tap}} side effect operator to log values. A less manually invasive way would be preferable as stated.

@@

rxfiddle proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time. They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.

>>>>

We organized informal interviews which allowed us to gain first insights in how software engineers work with RP in their daily jobs. We talked to six engineers and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.

Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications. By being the creator of a visualizer for RxJS \mintinline{TypeScript}{Observable}s, the fifth interviewee was a proficient RxJS user. Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.

All of the interviewees referred to RP with an overall positive experience. We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions. Hence a big strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL. The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.

It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often. E.g. when using ngrx for state management, ``the framework takes care that you do not have to interact with Observables on your own, but leverages greatly on their potential behind the scenes''. As soon as our interviewees had to extend built-in functionalities with their own features, e.g. a new effect\footnote{\url{https://ngrx.io/guide/effects}}, all of them valued the possibility to interact with underlying \mintinline{TypeScript}{Observable}s highly.

When asked explicitly about what they dislike the most about RP, all interview partners with no exception emphasized the debugging process of an RP application as a big nuisance. The fact that our interviewees remembered the search for a bug as something negative did not surprise us, hence a bug is usually something negative afflicted. What did alarm us though was the fact that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' were prevalent and gave insight in \emph{why} our interview partners dislike RP debugging in particular. Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless. Further, one single partner mentioned marble diagrams as very valuable in order to understand how an RxJS \mintinline{TypeScript}{Observable} works, no matter if during development or when debugging code. Marble diagrams are a way to visualize and describe the value-emitting behavior of an \mintinline{TypeScript}{Observable} over time. They are accepted throughout the ReactiveX community for documentation\cite{marblediagrams} and testing reasons\cite{marbletesting}.

@@

rxfiddle proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.
